Embedding Intuitionistic into Classical Logic

### 2022-07-14 CSL/lipics-v2021-EIICL.tex
Unified notation for prop sem. and pred sem. definition

Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and ILTP~\cite{iltp}, we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover} utilize classical logic. This difference can be partially explained by fundamental differences between the logics. First of all determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}. A further advantage of classical logic is the existence of calculi that are particularly suitable for automation like superposition~\cite{bachmair2001resolution} which rely on the existence of convenient normal forms such as CNF and the duality between validity and satisfiability, i.e. to show validity they show unsatisfiability of the negation. While some kind of normal form is of course be obtainable for intuitionistic logic, albeit more complex, crucially the duality does not hold, e.g. $\neg(A\vee \neg A)$ is not satisfiable while $A\vee\neg A$ is also not a theorem. Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly constructing a cut-free proof from the theorem by applying the inverse of rules from some proof calculus, enhanced by search strategies such as focussing and polarization. This strategy generally leads to a much more complex search and is therefore difficult to implement efficiently. Finally in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, this is particularly true for the propositional case, i.e. SAT-solvers.

@@

Before proceeding towards the results let us quickly fix our notation and recapitulate the semantics for intuitionistic logic.

@@

\begin{definition}
	A \textbf{valuation} $v$ is a function that assigns to each propositional variable $A$ a \textbf{truth value} $v(A)\in\{0, 1\}$. We extend $v$ to a function $v^*$ on formulas:
	\begin{itemize}
		\item $v^*(\bot) = 0$
		\item $v^*(A) = v(A)$ for each propositional variable $A$.
		\item $v^*(\varphi\wedge\psi) = \min(v^*(\varphi), v^*(\psi))$
		\item $v^*(\varphi\vee\psi) = \max(v^*(\varphi), v^*(\psi))$
		\item $v^*(\varphi\to \psi) = \max(1 - v^*(\varphi), v^*(\psi))$
	\end{itemize}
	$v$ is a \textbf{model} for $\varphi$ if $v(\varphi) = 1$. We denote the set of valid formulas with CPC.
\end{definition}

One notable property of classical logic is that a formula $\varphi$ is valid if and only if its negation, i.e. $\neg\varphi$ is not satisfiable. The same does not hold true for intuitionistic logic as we shall see.

\begin{definition}
	A \textbf{Kripke structure} $\mathcal K = (W, (v_w)_{w\in W})$ consists of a partially ordered set $W$ of nodes an a family of valuations $(v_w)_{w\in W}$ such that for $u\leq w$ we have $v_u(A)\leq v_u(A)$ for every propositional variable $A$, this is called the \textbf{persistency condition}. We extend the valuations $v_u$ slightly differently to $v_u^\#$ than before, i.e. have
	\begin{itemize}
		\item $v_u^\#(\bot) = 0$
		\item $v_u^\#(\varphi) = v_u(A)$ if $\varphi$ is atomic.
		\item $v_u^\#(\varphi\wedge\psi) = \min(v_u^\#(\varphi), v_u^\#(\psi))$
		\item $v_u^\#(\varphi\vee\psi) = \max(v_u^\#(\varphi), v_u^\#(\psi))$
		\item $v_u^\#(\varphi\to \psi) = \min\{\max(1 - v_w^\#(\varphi), v_w^\#(\psi))\:|\:w\geq u\}$
	\end{itemize}
	That is $v_u^\#(\varphi\to \psi) = 1$ if at every world $w\geq u$ we have $v_w^\#(\varphi) = 0$ or $v_w^\#(\psi) = 1$. We say that $\varphi$ is satisfied at a node $u$ if $v_u^\#(\varphi) = 1$ and write $u\models\varphi$. If a formula is satisfied at every node then $\mathcal K$ is a \textbf{model} for $\varphi$. We denote the set of valid formulas with IPC.
\end{definition}
There are many classical tautologies which are not intuitionistically valid
\begin{example}\label{LEMcounterexample}
	Consider the famous law of excluded middle $A\vee\neg A$. Recall that $\neg A := A\to \bot$. As a counter-model consider a Kripke structure with two nodes $u\leq w$ and $v_u(A) = 0, v_w(A) = 1$. Then $u\not\models A$ and $u\not\models \neg A$, i.e. $u\not\models A\vee\neg A$. 
\end{example}
In classical logic a proven strategy for showing that a formula valid is to consider its negation and show that it is not satisfiable. This is insufficient for intuitionistic logic, e.g. $A\vee\neg A$ is not a intuitionistic tautology but its negation is not satisfiable either.

@@

\begin{definition}
	A $\Sigma$-\textit{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$ and a family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$
	\begin{itemize}
		\item $M_u\subseteq M_w$.
		\item $f^{I_w}|_{M_u} = f^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
		\item $R^{I_w}|_{M_u} = R^{I_u}$.
	\end{itemize}
	Then we define a model relation between worlds $u\in W$ and variable assignments $v$ (targeting $M_u$) and formulas $\varphi$ as follows:
	\begin{itemize}
		\item $u, v\not\models\bot$
		\item $u, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I_u, v},\dots,t_n^{I_u, v})\in R^{I_u}$.
		\item $u, v\models s = t$ iff $s^{I_u, v} = t^{I_u, v}$.
		\item $u, v\models \varphi\wedge \psi$ iff $u, v\models\varphi$ and $u, v\models\psi$.
		\item $u, v\models \varphi\vee\psi$ iff $u, v\models\varphi$ or $u, v\models\psi$.
		\item $u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $w, v\not\models\varphi$ or $w, v\models\psi$.
		\item $u, v\models\exists x\varphi$ iff there exists $m\in M_u$ such that $u, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
		\item $u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $w, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
	\end{itemize}
	$\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. We denote the set of valid formulas with IQC.
\end{definition}
In addition to the propositional tautologies there are now sentences involving quantifiers which are classically valid but not intuitionistically.
Consider for instance the classical tautology $\varphi$ $$\neg\forall x A(x)\to \exists x \neg A(x)$$Analogous remarks with regard to validity and satisfiability hold as in the propositional case.

@@

\begin{example}
	Consider
	$$\mathcal S = \{(A\to B)\to C, (B\to A)\to D, (A\wedge B)\to \bot, (C\vee D)\to E\}$$and $\varphi = \bigwedge S\to E$ which has the Kripke counter-model
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right)   [right=of center] {$B, C$};
			\node[squarednode]      (left)    [left=of center] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left.south east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right.south west);			
		\end{tikzpicture}
	\end{center}
	\vspace*{-.3cm}where at each node the true propositions are indicated.\pagebreak

>>>>

Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and ILTP~\cite{iltp}, we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover} utilize classical logic. This difference can be partially explained by fundamental differences between the logics. First of all determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}. A further advantage of classical logic is the existence of calculi that are particularly suitable for automation like superposition~\cite{bachmair2001resolution} which rely on the existence of convenient normal forms such as CNF and the duality between validity and satisfiability, i.e. to show validity they show unsatisfiability of the negation. While some kind of normal form is of course be obtainable for intuitionistic logic, albeit more complex, crucially the duality does not hold, e.g. $\neg(A\vee \neg A)$ is not satisfiable while $A\vee\neg A$ is also not a theorem. Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly constructing a cut-free proof from the theorem by applying the inverse of rules from some proof calculus, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently. Finally in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, this is particularly true for the propositional case, i.e. SAT-solvers.

@@

Before proceeding towards the results let us quickly fix our notation and recapitulate the semantics for classical and more importantly intuitionistic logic.

@@

\begin{definition}
	A \emph{valuation} $v$ is a function that assigns to each propositional variable $A$ a \textbf{truth value} $v(A)\in\{0, 1\}$. We inductively define the model relation $\models$ between $v$ and formulas:
	\begin{itemize}
		\item $v\not\models \bot$
		\item $v\models A$ iff $v(A) = 1$ for each propositional variable $A$.
		\item $v\models \varphi\wedge\psi$ iff $v\models\varphi$ and $v\models\psi$.
		\item $v\models\varphi\vee\psi$ iff $v\models\varphi$ or $v\models\psi$.
		\item $v\models\varphi\to \psi$ iff $v\not\models\varphi$ and $v\models\psi$.
	\end{itemize}
	$v$ is a \textbf{model} for $\varphi$ if $v\models\varphi$. If every valuation is a model for $\varphi$ then we say $\varphi$ is \emph{valid}. We denote the set of valid formulas with \emph{CPC}.
\end{definition}

One notable property of classical logic is that a formula $\varphi$ is valid if and only if its negation $\neg\varphi := \varphi\to\bot$ is not satisfiable, i.e. there does not exist a model for it. The same does not hold true for intuitionistic logic as we shall see.

\begin{definition}
	A \emph{Kripke structure} $\mathcal K = (W, (v_w)_{w\in W})$ consists of a partially ordered set $W$ of nodes an a family of valuations $(v_w)_{w\in W}$ such that for $u\leq w$ we have $v_u(A)\leq v_u(A)$ for every propositional variable $A$, this is called the \emph{persistency condition}. The model relation between the $v_u$ and formulas $\varphi$ is defined as before, except in the case of implications, i.e. have
	\begin{itemize}
		\item $v_u\models\varphi\to \psi$ if for all $w\geq u$ we have $w\not\models\varphi$ or $u\models\psi$.
	\end{itemize}
	We say that $\varphi$ is satisfied at a node $u$ if $v_u\models\varphi$ and write $u\models\varphi$. If a formula is satisfied at every node then $\mathcal K$ is a \textbf{model} for $\varphi$. A formula is valid if every Kripke structure is a model of it. We denote the set of valid formulas with IPC.
\end{definition}
There are many classical tautologies which are not intuitionistically valid
\begin{example}\label{LEMcounterexample}
	Consider the famous law of excluded middle $A\vee\neg A$. Recall that $\neg A := A\to \bot$. As a counter-model consider a Kripke structure with two nodes $u\leq w$ and $v_u(A) = 0, v_w(A) = 1$. Then $u\not\models A$ and $u\not\models \neg A$, i.e. $u\not\models A\vee\neg A$. 
\end{example}
In classical logic a proven strategy for showing that a formula valid is to consider its negation and show that it is not satisfiable. This is insufficient for intuitionistic logic, e.g. $A\vee\neg A$ is not a intuitionistic tautology but its negation is not satisfiable either.

@@

\begin{definition}
	A $\Sigma$-\emph{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$ and a family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$
	\begin{itemize}
		\item $M_u\subseteq M_w$.
		\item $f^{I_w}|_{M_u} = f^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
		\item $R^{I_w}|_{M_u} = R^{I_u}$.
	\end{itemize}
	Then we define a model relation between worlds $u\in W$ and variable assignments $v$ (targeting $M_u$) and formulas $\varphi$ as follows:
	\begin{itemize}
		\item $u, v\not\models\bot$
		\item $u, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I_u, v},\dots,t_n^{I_u, v})\in R^{I_u}$.
		\item $u, v\models s = t$ iff $s^{I_u, v} = t^{I_u, v}$.
		\item $u, v\models \varphi\wedge \psi$ iff $u, v\models\varphi$ and $u, v\models\psi$.
		\item $u, v\models \varphi\vee\psi$ iff $u, v\models\varphi$ or $u, v\models\psi$.
		\item $u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $w, v\not\models\varphi$ or $w, v\models\psi$.
		\item $u, v\models\exists x\varphi$ iff there exists $m\in M_u$ such that $u, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
		\item $u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $w, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
	\end{itemize}
	$\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. We denote the set of valid formulas with IQC.
\end{definition}
In addition to the propositional tautologies there are now sentences involving quantifiers which are classically valid but not intuitionistically.
Consider for instance the classical tautology $\varphi$ $$\neg\forall x A(x)\to \exists x \neg A(x)$$Analogous remarks with regard to validity and satisfiability hold as in the propositional case.

@@

\begin{example}
	Consider
	$$\mathcal S = \{(A\to B)\to C, (B\to A)\to D, (A\wedge B)\to \bot, (C\vee D)\to E\}$$and $\varphi = \bigwedge S\to E$ which has the Kripke counter-model
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right)   [right=of center] {$B, C$};
			\node[squarednode]      (left)    [left=of center] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left.south east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right.south west);			
		\end{tikzpicture}
	\end{center}
	\vspace*{-.3cm}where at each node the true propositions are indicated.
